Handling Recursion in Generic Programming Using Closed Type Families

### 2018-07-18 lncs-paper.tex
Polymorphic recursion

\newcommand{\Ra}{$\Rightarrow$\s}
\newcommand{\ra}{$\rightarrow$\s}
\newcommand{\fa}{$\forall$}
\newcommand{\tteq}{{\scriptsize$\thicksim$}\s}
\newcommand{\comp}{{\tt>\nhs>\nhs>}}
\newcommand{\kleisli}{{\tt>\nhs=\nhs>}}
\newcommand{\comps}{\comp\s}
\newcommand{\kleislis}{\kleisli\s}
\newcommand{\ann}{:\nhs:\s}

@@

We feel both existing approaches unsatisfactory and make the following contributions.
\begin{itemize}
\item We describe the problem with the current approach of SOP in detail (Section~\ref{sec:sop-problem}).
\item We introduce an idiom that overcomes the problem. The approach avoids both, the use of overlapping instances and changing a generic representation. (Section~\ref{sec:handling-recursion})
% -- 
%\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
% -- this doesn't sound like a contribution
\item We evaluate our approach through the development of a larger-scale use case---the generic zipper. The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).
\item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations, which necessarily  arise in practice as a consequence of absence of information about recursion points. An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{sec:gshow}.
\end{itemize}

@@

We illustrate the problem through a short example. The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure. We reimplement this function using the SOP view:
\texttt{
\vs
\indent subterms \ann Generic a \Ra a \ra [a]\\
\indent subterms t = subtermsNS (unSOP \$ from t)
\vs
\indent subtermsNS \ann NS (NP I) xss \ra [a]\\
\indent subtermsNS (S ns) = subtermsNS ns\\
\indent subtermsNS (Z np) = subtermsNP np
\vs
\indent subtermsNP \ann \fa a xs. NP I xs \ra [a]\\
\indent subtermsNP p (I y :* ys)\\
\indent\s\s | typeEq @a y \hs\hs\ths= witnessEq y : subtermsNP ys\\
\indent\s\s | otherwise\s\s\s\hs\hs\ths = subtermsNP ys\\
\indent subtermsNP \_ Nil = []
\vs
}
The function \texttt{subterms} translates the term to its representation, unwrapping the sum of products from \texttt{SOP}, and passes that to the auxiliary function \texttt{subtermsNS}. The latter merely traverses the sum and, once reaches the product,  passes it further to \texttt{subtermsNP}.

@@

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms} task. We give another example showing how to adapt our idiom to different scenarios in the following subsection.

\subsection{Generic show}
\label{sec:gshow}

@@

@@

The subtree in a focus is wrapped by the \texttt{Focus} datatype. 
The wrapper incapsulates the proofs about a number of important properties of a focus.
\texttt{
\vs
\indent data Focus (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\s\s\s (c \ann * \ra Constraint) where\\
\indent\s\s Focus \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
\indent\ind\ind\Ra b \ra Focus r a fam c
\vs
}
Existential type variable \texttt{b} represents the type of a focus. We apply a number of predicates to \texttt{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus. Firstly, the type of a focus should have the \texttt{Generic} representation. Secondly, it should live \texttt{In} the list of types we are going to visit. Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \texttt{ZipperI} predicate. In particular, the predicate ensures that \texttt{a} is the type of the parent for the focus in the structure under consideration.

@@

Finally, a context frame has two representations: as a type synonym and as a newtype wrapper
\texttt{
\vs
\indent type CtxCode  fam a = ToContext 'F fam (Code a)
\vs
\indent newtype Context fam a = Ctx \{ctx \ann SOP I (CtxCode fam a)\}
\vs
}
The wrapper allows GHC to perform type inference where it is doomed to fail with a plain type synonym. On the other hand, the \texttt{CtxCode} type comes handy in constraints applied to the generic zipper interface functions.

@@

>>>>

\newcommand{\Ra}{$\Rightarrow$\s}
\newcommand{\ra}{$\rightarrow$\s}
\newcommand{\fa}{$\forall$}
\newcommand{\tteq}{{\scriptsize$\thicksim$}\s}
\newcommand{\comp}{{\tt>\nhs>\nhs>}}
\newcommand{\kleisli}{{\tt>\nhs=\nhs>}}
\newcommand{\comps}{\comp\s}
\newcommand{\kleislis}{\kleisli\s}
\newcommand{\ann}{:\nhs:\s}
\newcommand{\ds}{.\nhs.\nhs.}

@@

We feel both existing approaches unsatisfactory and make the following contributions.
\begin{itemize}
\item We describe the problem with the current approach of SOP in detail (Section~\ref{sec:sop-problem}).
\item We introduce an idiom that overcomes the problem. The approach avoids both, the use of overlapping instances and changing a generic representation (Section~\ref{sec:handling-recursion}).
% -- 
%\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
% -- this doesn't sound like a contribution
\item We evaluate our approach through the development of a larger-scale use case---the generic zipper. The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).
\item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations, which necessarily  arise in practice as a consequence of absence of information about recursion points. An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{subsec:gshow}.
\end{itemize}

@@

We illustrate the problem through a short example. The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure. We reimplement this function using the SOP view:
\texttt{
\vs
\indent subterms \ann Generic a \Ra a \ra [a]\\
\indent subterms t = subtermsNS \$ unSOP \$ from t
\vs
\indent subtermsNS \ann NS (NP I) xss \ra [a]\\
\indent subtermsNS (S ns) = subtermsNS ns\\
\indent subtermsNS (Z np) = subtermsNP np
\vs
\indent subtermsNP \ann \fa a xs. NP I xs \ra [a]\\
\indent subtermsNP p (I y :* ys)\\
\indent\s\s | typeEq @a y \hs\hs\ths= witnessEq y : subtermsNP ys\\
\indent\s\s | otherwise\s\s\s\hs\hs\ths = subtermsNP ys\\
\indent subtermsNP \_ Nil = []
\vs
}
The function \texttt{subterms} translates the term to its representation, unwrapping the sum of products from \texttt{SOP}, and passes that to the auxiliary function \texttt{subtermsNS}. The latter merely traverses the sum and, once reaches the product,  passes it further to \texttt{subtermsNP}.

@@

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms}' task. We give another example showing how to adapt our idiom to different scenarios in the following subsection.

\subsection{Generic show}
\label{subsec:gshow}

@@

\section{Polymorphic recursion}
\label{sec:polymorphic-recursion}

The solution, described in the previous section, is applicable to a range of datatypes that are \emph{monomorphically recursive}. Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g. \texttt{Tree~a} from Section~\ref{sec:sop-view}). We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent. It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested datatypes.

Assume we have a polymorphically recursive datatype \texttt{PolyRec~a} defined in terms of a type family \texttt{Poly}:
\texttt{
\vs
\indent data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
\vs
\indent type family Poly a where\\
\indent \ind  Poly Bool = Char\\
\indent \ind  Poly Char = Bool\\
\indent \ind  Poly a\ind = a
\vspace{0.2cm}
}

For managing polymorphic recursion in this datatype, we can write an analogue of the \texttt{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types. Since any datatype with type parameters \texttt{f a b c \ds} has kind \texttt{* \ra (* \ra (* \ra \ds))}, a \texttt{PolyEq} type family can be defined thus:
\texttt{
\vs
\indent type family PolyEq (a \ann k) (x \ann k) \ann Bool where\\
\indent \ind  PolyEq (f a) (g b) = PolyEq f g\\
\indent \ind  PolyEq a\ind\s a\ind\s = 'True\\
\indent \ind  PolyEq a\ind\s x\ind\s = 'False
\vspace{0.2cm}
}

The function \texttt{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \texttt{Equal}. The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type:
\texttt{
\vs
\indent instance Show \s b \Ra CaseShow 'False a b where\\
\indent \ind    caseShow' \s\s = show\\
\indent instance GShow b \Ra CaseShow 'True \s a b where\\
\indent \ind    caseShow' t = "(" ++ gshow t ++ ")"
\vspace{0.2cm}
}

Unfortunately, this approach, albeit working well for datatypes defined like \texttt{PolyRec~a}, becomes unsuitable for nested datatypes, such as one below:
\texttt{
\vs
\indent data Nested a = Epsilon | Nest a (Nested [a])
\vs
}
The culprit is the constraint \texttt{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \texttt{GShow b} instance context above.

@@

The subtree in focus is wrapped by the \texttt{Focus} datatype. 
The wrapper incapsulates the proofs about a number of important properties of a focus.
\texttt{
\vs
\indent data Focus (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\s\s\s (c \ann * \ra Constraint) where\\
\indent\s\s Focus \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
\indent\ind\ind\Ra b \ra Focus r a fam c
\vs
}
Existential type variable \texttt{b} represents the type of a focus. We apply a number of predicates to \texttt{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus. Firstly, the type of a focus should have the \texttt{Generic} representation. Secondly, it should live \texttt{In} the list of types we are going to visit. Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \texttt{ZipperI} predicate. In particular, the predicate ensures that \texttt{a} is the type of the parent for the focus in the structure under consideration.

@@

Finally, a context frame has two representations: as a type synonym and as a newtype wrapper:
\texttt{
\vs
\indent type CtxCode  fam a = ToContext 'F fam (Code a)
\vs
\indent newtype Context fam a = Ctx \{ctx \ann SOP I (CtxCode fam a)\}
\vs
}
The wrapper allows GHC to perform type inference where it is doomed to fail with a plain type synonym. On the other hand, the \texttt{CtxCode} type comes handy in constraints applied to the generic zipper interface functions.

@@

\section*{Acknowledgments}
\label{sec:acks}

We are grateful to Julia Belyakova, who helped to proof-read certain parts of the paper, 
and to the participants of the Seminar on Programming Languages and Compilers at 
I.I.~Vorovich Institute of Mathematics, Mechanics, and Computer Science (Southern Federal
University, Russia), where we presented partial results of the work and got valuable feedback.

This project has received funding from the European Research Council (ERC) under the
European Union's Horizon 2020 research and innovation programme (grant agreement No 695412).
